---
title: "Let's Analyse Penguin Data with R"
author: "Adnan Ali Syed"
date: 2026-09-28
---

# Introduction

In this post, we will explore the Palmer Penguins dataset using R.
We will compare the body mass and flipper length of the three penguin species.

The data comes from the [Palmer Penguins dataset](https://allisonhorst.github.io/palmerpenguins/), 
collected by Palmer Station Antarctica LTER. The dataset is available directly
through the `palmerpenguins` R package.

# Load and clean the data

```{r}
#| message: false
#| warning: false

library(palmerpenguins)

penguins_clean <- na.omit(
  penguins[, c(
    "species",
    "island",
    "bill_length_mm",
    "flipper_length_mm",
    "body_mass_g"
  )]
)

head(penguins_clean)
```

I selected the variables needed for this analysis and removed rows containing
missing values.

# Calculate average measurements by species

```{r}
species_summary <- aggregate(
  cbind(body_mass_g, flipper_length_mm) ~ species,
  data = penguins_clean,
  FUN = mean
)

species_summary
```

The table shows that the three species differ in both average body mass and
average flipper length.

# Count the number of penguins in each species

```{r}
species_counts <- table(penguins_clean$species)

species_counts
```

This tells us how many observations from each species remain after removing
missing values.

# Compare body mass visually

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of Palmer penguins by species."

boxplot(
  body_mass_g ~ species,
  data = penguins_clean,
  xlab = "Species",
  ylab = "Body mass (g)"
)
```

The boxplot shows noticeable differences in body mass between the species.
Gentoo penguins generally have higher body masses than Adelie and Chinstrap
penguins.

# Conclusion

The R analysis gives a similar result to the Python analysis. The penguin
species have different physical characteristics, and Gentoo penguins are
generally heavier and have longer flippers than the other two species.