---
title: "Exploring the Gapminder Dataset with R"
author: "Adnan Ali Syed"
date: 2026-09-28
---

# Introduction

In this post, I will explore the Gapminder dataset using R. The goal is to
look at differences in life expectancy and GDP per capita across continents,
and then look at how life expectancy has changed over time for one country.

The data comes from the
[Gapminder dataset](https://www.gapminder.org/data/) and is available in R
through the `gapminder` package.

# Load the packages

```{r}
#| message: false
#| warning: false

library(gapminder)
library(dplyr)
library(ggplot2)
```

The `gapminder` dataset contains information about countries across several
years, including population, life expectancy, and GDP per capita.

# Look at the most recent year

For the first part of the analysis, I will focus on the year 2007 and select
only the variables that are useful for this comparison.

```{r}
gapminder_2007 <- gapminder |>
  filter(year == 2007) |>
  select(country, continent, lifeExp, gdpPercap, pop)

head(gapminder_2007)
```

The resulting dataset contains one observation for each country in 2007.

# Average life expectancy by continent

Next, I will group the countries by continent and calculate the average life
expectancy for each continent.

```{r}
continent_summary <- gapminder_2007 |>
  group_by(continent) |>
  summarise(
    average_life_expectancy = mean(lifeExp),
    average_gdp_per_capita = mean(gdpPercap)
  ) |>
  arrange(desc(average_life_expectancy))

continent_summary
```

This table allows us to compare the average life expectancy and GDP per capita
across continents. The continents with higher average GDP per capita also tend
to have higher average life expectancy in this dataset.

# GDP per capita and life expectancy

We can also look at the relationship between GDP per capita and life expectancy
for individual countries in 2007.

```{r}
#| label: fig-gdp-life-expectancy
#| fig-cap: "GDP per capita and life expectancy for countries in 2007."

ggplot(
  data = gapminder_2007,
  mapping = aes(
    x = gdpPercap,
    y = lifeExp,
    colour = continent
  )
) +
  geom_point(size = 2) +
  labs(
    x = "GDP per Capita",
    y = "Life Expectancy",
    colour = "Continent"
  ) +
  theme_bw()
```

The scatter plot shows that countries with higher GDP per capita generally
have higher life expectancy. There are also noticeable differences between
continents.

# Life expectancy in Canada over time

Finally, I will look at how life expectancy changed over time in Canada.

```{r}
canada_data <- gapminder |>
  filter(country == "Canada")

canada_data
```

The filtered dataset contains the observations for Canada from 1952 to 2007.

```{r}
#| label: fig-canada-life-expectancy
#| fig-cap: "Life expectancy in Canada from 1952 to 2007."

ggplot(
  data = canada_data,
  mapping = aes(
    x = year,
    y = lifeExp
  )
) +
  geom_line() +
  geom_point() +
  labs(
    x = "Year",
    y = "Life Expectancy"
  ) +
  theme_bw()
```

The graph shows that life expectancy in Canada increased steadily over the
period represented in the Gapminder dataset.

# Conclusion

This analysis used the Gapminder dataset to explore life expectancy and GDP per
capita. The 2007 data shows differences between continents and suggests a
positive relationship between GDP per capita and life expectancy. Looking at
Canada separately also shows that life expectancy increased over time.